Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-02-11 manuscript/climate_sensitivity_draft.Rmd
worked on intro

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to be more affected by drought. Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016), although there are exceptions [@zuleta_drought-induced_2017]. It has yet to be resolved which of several potential underlying mechanisms drive this pattern. First, taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009]. To meet this challenge, hydraulic traits vary in relation to tree height within indivdiuals {**REFS**}, within species {**REFS**}, and across species [@liu_hydraulic_2019]. Despite the tendency for hydrualic traits to become more conservative with height on an individual tree {**REFS**}, the need for greater hydraulic efficiency (*i.e.,* larger xylem conduits) in tall trees [@liu_hydraulic_2019]{**REFS**} may leave them more vulnerable during drought {**REFS-- Mark Olsen--mentioned by Lawren**}. Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**). Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer a potential advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure. A final mechanism that could mediate tree size-related responses to drought is how species, and their associated hydraulic traits, are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019]. Within a forest community, greater drought vulnerability of larger trees could occur if the larger size classes were dominated by species less adapted to handle drought. Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.  {**ask Lawren to check revisions to this paragraph**}

**the following text needs to be integrated into the paragraph above (KAT)**
Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018]. Vertical gradients in stem and leaf traits-including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]-make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges [@olson_plant_2018]. Trade-offs can also occur, whereby traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms, for example in ponderosa pine [@roskilly_conflicting_2019].

Debates have also arisen regarding the traits that would influence tree growth responses to drought. Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017]. However, in other cases these traits have failed to explain drought responses (**REFS**), and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017] but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, their role may be due to indirect correlations with other traits within life-history strategies [@hoffmann_hydraulic_2011]. Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. Hydraulic traits including water potentials at which percent loss conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019]. These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

The long time frame captured in tree-ring data enables us to address the question of whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits. Tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains. While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment. The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on the most extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016). Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals on the order of one to two decades - remains limited. There is evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019], so the influence of tree size may be less in weaker--but more common--droughts. Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics, and the extent to which their influence is consistent across droughts.

@@

Our analysis indicates that tree height has a stronger influence on drought response than does canopy position (Tables 1,4,5). This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@couvreur_water_2018; @roskilly_conflicting_2019] **REFS**. However, this result must be interpeted with some caution, given that colinearity between the two variables (Fig. **2d**) makes it impossible to confidently partition causality. Taller trees are more likely to be in dominant canopy positions (Fig. **2d**) and, largely as a consequence of their position relative to others, face different microenvironments (Fig. **2a-b**). Even under non-drought conditions, evaporative demand and maximum leaf teamperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur extra stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. However, some decoupling between height and canopy position is introduced by the configuration of neighboring trees (Fig. 2d) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Tables 1,4,5).

>>>>

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to be more affected by drought. Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016). It has yet to be resolved which of several potential underlying mechanisms drive this pattern. First, taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018]. Vertical gradients in stem and leaf traits-including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]-make it biophysically possible for trees to become tall [@couvreur_water_2018]. Meanwhile, tall trees require greater hydraulic efficiency, such that xylem conduit diameter increases with tree height within and across species [@olson_plant_2018; @liu_hydraulic_2019], making large trees more vulnerable to embolism during drought [@olson_plant_2018]. Traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019]. Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity (e.g., @koike_leaf_2001; @kunert_revised_2017). Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer a potential advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure. A final mechanism that could mediate tree size-related responses to drought is how species, and their associated hydraulic traits, are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019]. Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.

Debates have also arisen regarding the traits that would influence tree growth responses to drought. Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017]. However, in other cases these traits have failed to explain drought responses, and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017] but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, their role may be due to indirect correlations with other traits within life-history strategies [@hoffmann_hydraulic_2011]. Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. Hydraulic traits including water potentials at which percent loss conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019]. These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

The long time frame captured in tree-ring data enables us to address the question of whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits. Tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains. While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., @lloret_components_2011;  @dorangeville_drought_2018), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment. The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on the most extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016). Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals on the order of one to two decades - remains limited. There is evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019], so the influence of tree size may be less in weaker--but more common--droughts. Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics, and the extent to which their influence is consistent across droughts.

@@

Our analysis indicates that tree height has a stronger influence on drought response than does canopy position (Tables 1,4,5). This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. However, this result must be interpeted with some caution, given that colinearity between the two variables (Fig. **2d**) makes it impossible to confidently partition causality. Taller trees are more likely to be in dominant canopy positions (Fig. **2d**) and, largely as a consequence of their position relative to others, face different microenvironments (Fig. **2a-b**). Even under non-drought conditions, evaporative demand and maximum leaf teamperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur extra stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. However, some decoupling between height and canopy position is introduced by the configuration of neighboring trees (Fig. 2d) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Tables 1,4,5).
